---
title: "PPI Network Analysis"
description: |
 This script constructs a protein-protein interaction network using the STRING database and visualizes the network via several plots.
bibliography: astrocyte-review-bibliography.bib
csl: https://www.zotero.org/styles/elsevier-vancouver
author:
  - first_name: "Ayush"
    last_name: "Noori"
    url: https://www.github.com/ayushnoori
    affiliation: Massachusetts General Hospital
    affiliation_url: https://www.serranopozolab.com
    orcid_id: 0000-0003-1420-1236
output:
  distill::distill_article:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Dependencies

Load requisite packages and define directories. Note that the GitHub package [`mattflor/chorddiag`](https://github.com/mattflor/chorddiag) is used to create interactive chord diagrams using the Javascript visualization library D3. This package can be downloaded via `devtools::install_github("mattflor/chorddiag")`.

This script also uses my personal utilities package `brainstorm`, which can be downloaded via `devtools::install_github("ayushnoori/brainstorm")`.

```{r load-packages, message=FALSE, warning=FALSE}

# data manipulation
library(data.table)
library(purrr)
library(magrittr)

# fast file system operations
library(fs)

# access STRING API
library(httr)

# base visualization
library(ggplot2)
library(RColorBrewer)

# graph libraries
library(igraph)
library(ggraph)
library(graphlayouts)

# heatmap
library(ComplexHeatmap)

# chord diagrams
library(circlize)
library(chorddiag)

# interactive network
library(visNetwork)

# utility functions
library(brainstorm)

```

Note that directories are relative to the R project path.

```{r define-directores}

# set directories
ddir = "Data"
dir1 = file.path("Results", "1 - STRING PPI Network")

```

# Query STRING API

Read data, then count the number of input proteins.

```{r read-data}

# read data
dat = fread(file.path(ddir, "ADRA Protein Set.csv"), encoding = "UTF-8")
nrow(dat) %>% paste0("Input Proteins: ", .) %>% message

# show data
show_table(head(dat, 20))

```

[STRING](https://string-db.org/) is a database of known and predicted protein-protein interactions which allow the creation of functional protein association networks [@szklarczyk_string_2019].

Map from gene symbols to STRING IDs using the STRING API. Proteins whose preferred IDs differ from their query symbol are later replaced before the network is constructed [@the_uniprot_consortium_uniprot_2021; @tweedie_genenamesorg_2021]. As expected, the immunoglobulins are excluded as they are not represented in the STRING database.

```{r map-symbols}

root_api = "https://version-11-0b.string-db.org/api"

# construct query URL
map_symbols = list(identifiers = paste(dat[, Symbol], collapse="%0d"),
                   species = "9606", echo_query = "1", caller_identity = "SerranoPozoLab")

# complete API call to map IDs
id_request = httr::POST(url = paste0(root_api, "/tsv/get_string_ids"), body = map_symbols)
ids = httr::content(id_request, as = "text", encoding = "UTF-8") %>% fread()

# check for multiple mappings
ids[, any(duplicated(queryIndex))] %>% paste0("Multiple Mappings: ", ., "\n") %>% message()

# check for duplicate mappings
diff = ids[queryItem != preferredName]
diff[, paste(queryItem, preferredName, sep = " --> ")] %>%
  c("Different Mappings: ", .) %>% paste0("\n") %>% message()

# count total proteins with mappings
nrow(ids) %>% paste0("Proteins with Mappings: ", ., "/", nrow(dat), "\n") %>% message()

# print excluded proteins
dat[!(Symbol %in% ids$queryItem), paste(Symbol, collapse = ", ")] %>%
  paste0("Excluded Markers: ", .) %>% message()

```

Construct the network using another API call. Note that, as above, we use a `POST` request rather than the simpler `GET` to circumvent the character limit of the latter. KIF21B is excluded as it is an isolated node (does not have any connections).

```{r get-network}

# construct query URL
get_network = list(identifiers = paste(ids[, stringId], collapse="%0d"),
                   species = "9606", echo_query = "1", caller_identity = "SerranoPozoLab")

# complete API call to retrieve network
network_request = httr::POST(url = paste0(root_api, "/tsv/network"), body = get_network)
network = httr::content(network_request, as = "text", encoding = "UTF-8") %>% fread()

# count total included proteins
network[, c(preferredName_A, preferredName_B)] %>% uniqueN() %>%
  paste0("Proteins Included in Network: ", ., "/", nrow(dat), "\n") %>% message()

# print excluded proteins
network[, c(preferredName_A, preferredName_B)] %>% 
  c(diff$queryItem, .) %>%
  { dat[!(Symbol %in% .), paste(Symbol, collapse = ", ")] } %>%
  paste0("Excluded Markers: ", .) %>% message()

```

# Construct PPI Network

Use `igraph` functions to create the network. Proteins whose preferred IDs differ from their query symbol are replaced before the network is constructed. Note that the complete `dat` object, including the markers IGHA1, IGHG1, IGHM, and KIF21B (which have no connections), is passed to the `graph_from_data_frame` call; hence, while these nodes will be included in the vertex list of the network object, they will have a centrality of 0 and will not be present when visualizing the network.

```{r construct-network}

# remove unneeded columns and duplicate rows
network[, c("stringId_A", "stringId_B", "ncbiTaxonId") := NULL]
network = unique(network)

# replace with correct query symbols
replace_diff = function(net, r, pname, qname) { net[get(r) == pname, (r) := qname] }
pwalk(diff[, .(preferredName, queryItem)], ~replace_diff(network, "preferredName_A", .x, .y))
pwalk(diff[, .(preferredName, queryItem)], ~replace_diff(network, "preferredName_B", .x, .y))
setnames(network, old = 1:2, c("nodeA", "nodeB"))

# construct network
setcolorder(dat, "Symbol")
net = graph_from_data_frame(d = network, vertices = dat, directed = FALSE)

# show network
show_table(head(network, 20))

```

Below is a key of column names which represent evidence for interaction in the STRING database.

| Name   | Definition                 |
|--------|----------------------------|
| score  | combined score             |
| nscore | gene neighborhood score    |
| fscore | gene fusion score          |
| pscore | phylogenetic profile score |
| ascore | coexpression score         |
| escore | experimental score         |
| dscore | database score             |
| tscore | text mining score          |

# Calculate Assortativity Coefficient

The assortativity coefficient is positive is similar vertices - based on some external property, in this case, we use functional category - tend to connect to each other, and negative otherwise.

```{r calculate-assortativity}

# calculate assortativity
assort = V(net)$Group %>% as.factor() %>% as.integer() %>% assortativity_nominal(net, ., directed = FALSE)
message(paste("Assortativity Coefficient:", assort))

```

# Define Color Palette

First, define the color palette for all subsequent plots.

```{r create-palette}

# define color palette
cols = colorRampPalette(c("#B4436C", "#F2BAC9", "#F7A278", "#FCEFB4", "#C8E9A0", "#6DD3CE", "#91A6FF", "#E5C2FF", "#B49082", "#ABB8C4"))(uniqueN(dat[, Group]))

# set names and order
names(cols) = dat[, Group] %>% unique() %>% .[order(.)]

# set alternating order for chord diagram
alt = dat[, .N, by = Group] %>% .[order(Group), N] %>% order()
alt = c(rbind(alt[1:9], rev(alt[10:18])))

# set order for chord diagram
# alt = dat[, .N, by = Group] %>% .[order(Group), N] %>% order()

```

# Plot Heatmap

Plot a heatmap of the adjacency matrix. Cluster rows and columns and annotate by functional group.

```{r plot-heatmap}

# extract adjacency matrix
hm_dat = as_adj(net, attr = "score", sparse = FALSE)
diag(hm_dat) = 1

# create heatmap annotations
hm_annos = dat[, Group]
hm_cols = list(Group = cols)

# create group labels, add line breaks
hm_grp = unique(hm_annos)
hm_grp[c(2, 3, 7, 8, 18)] = c("Blood-Brain\nBarrier", "Calcium\nHomeostasis", "Insulin\nSignaling", "Intracellular\nTrafficking", "Water/K+\nHomeostasis")

# top annotation
top_annos = HeatmapAnnotation(Group = anno_block(gp = gpar(fill = cols),
                                                 labels = hm_grp,
                                                 labels_gp = gpar(fontsize = 7, lineheight = 0.8)))

# left annotation
left_annos = rowAnnotation(Group = anno_block(gp = gpar(fill = cols),
                                              labels = hm_grp,
                                              labels_gp = gpar(fontsize = 7, lineheight = 0.8)))

# function to set outline of each functional group
group_outline = function(j, i, x, y, width, height, fill) {
  if(i[1] == j[1]) grid.rect(gp = gpar(lwd = 2, fill = "transparent"))
}

# function to establish color scale
color_scale = function(maxcol, val) { colorRamp2(c(0, 1), c("#F3F4F7", maxcol))(val) }

# function to set color of each cell, darker color is #9DA5BE
color_cell = function(j, i, x, y, width, height, fill) {
  if(hm_annos[i] == hm_annos[j]) {
    grid.rect(x = x, y = y, width = width, height = height,
              gp = gpar(col = NA, fill = color_scale(cols[hm_annos[i]], hm_dat[i, j])))
  } else {
    grid.rect(x = x, y = y, width = width, height = height,
             gp = gpar(col = NA, fill = color_scale("#525C7A", hm_dat[i, j])))
  }
}

# plot heatmap
hm = Heatmap(hm_dat,
             col = c("#F3F4F7", "red"),
             row_split = hm_annos, column_split = hm_annos,
             cluster_row_slices = FALSE, cluster_column_slices = FALSE,
             row_gap = unit(1, "mm"), column_gap = unit(1, "mm"), 
             column_title = NULL, row_title = NULL,
             top_annotation = top_annos, left_annotation = left_annos,
             # cell_fun = color_cell,
             layer_fun = group_outline,
             show_heatmap_legend = FALSE
             )

# save heatmap
pdf(file.path(dir1, "Network Heatmap.pdf"), width = 30, height = 30)
print(hm)
dev.off()

```

# Calculate Centrality Scores

Calculate the eigenvalue centrality scores for the entire graph. Then, remove isolated nodes for further visualization.

```{r calculate-centrality}

# calculate centrality
cent = eigen_centrality(net)$vector %>% data.table(Symbol = names(.), Centrality = .)

# add to vertex data
dat = dat %>% merge(cent, by = "Symbol") %>% .[order(-Centrality)]
vertex_attr(net, "Centrality") = cent[, Centrality]

# clean ID mapping
ids = ids[, .(queryItem, stringId, preferredName, annotation)]
setnames(ids, c("Symbol", "STRING.ID", "STRING.Symbol", "STRING.Annotation"))

# merge and save marker information
dat %>% merge(ids, by = "Symbol", all.x = TRUE, sort = FALSE) %>%
  fwrite(file.path(dir1, "STRING ADRA Annotations.csv"))

# remove isolated vertices from network
sub = delete_vertices(net, which(igraph::degree(net) <= 2))
paste0("Proteins Retained in Network: ", length(V(sub)), "/", nrow(dat)) %>% message()

```

# Plot Circos Graph

First, create a generic network plotting function which is called multiple times throughut the script.

```{r plot-network}

plot_network = function(net, layout, minmax, ..., arc = TRUE) {
  
  edge_geom = ifelse(arc, geom_edge_arc0, geom_edge_link0)
  
  p = ggraph(net, layout, ...) + 
    edge_geom(aes(width = escore), alpha = 0.4) +
    scale_edge_width(range = c(0.2, 0.9)) +
    geom_node_point(aes(color = Group, size = Centrality)) + scale_size(range = minmax) +
    scale_color_manual(values = cols) +
    theme_graph(fg_text_colour = "black", base_family = "Helvetica") + 
    guides(edge_width = FALSE, size = FALSE) +
    labs(color = "Group")
  
  return(p)
  
}

```

Next, create a function to plot a circos graph of the network, then plot the figure with and without labels. The labeling code is inspired by [this tutorial](https://www.r-graph-gallery.com/296-add-labels-to-circular-barplot) from the R Graph Gallery. Spaces are used as an efficient trick to keep labels from overlapping with the nodes. Although the labels are truncated in the figure, they can be recovered with Inkscape (or an analogous vector editing tool).

```{r plot-circos}

ngrp = uniqueN(vertex_attr(sub, "Group"))

plot_circos = function(net, minmax, fname, w, h, legend = FALSE) {
  
  p = plot_network(net, "linear", minmax, circular = TRUE, sort.by = Group)
  
  if(!legend) { p = p + theme(legend.position = "none") } else {
    
    p = p + # geom_node_label(aes(label = name), repel = TRUE, alpha = 0.8, segment.size = NA) +
      geom_node_text(aes(label = Label, angle = Angle, hjust = Hjust)) +
      theme(legend.title = element_text(face = "bold", size = 18, hjust = 0.5),
            legend.text = element_text(size = 14), legend.position = "bottom",
            plot.margin = unit(rep(-1, 4), "cm"))
  }
  
  ggsave(file.path(dir1, fname), p, width = w, height = h)
  
}

# generate angle for labels
md = sub %>%
  { data.table(Name = vertex_attr(., "name"), Angle = length(V(.))) } %>%
  .[, Angle := Angle %>% { (1:.[1] - 0.5)/. } %>% { 90 - (360*.) }] %>%
  .[, Hjust := ifelse(Angle < -90, 1, 0)] %>%
  .[, Label := ifelse(Angle < -90, paste0(Name, "      "), paste0("      ", Name))] %>%
  .[, Angle := Angle %>% ifelse(. < -90, . + 180, .)]

# assign to plot
vertex_attr(sub, "Angle") = md$Angle
vertex_attr(sub, "Hjust") = md$Hjust
vertex_attr(sub, "Label") = md$Label


# create and save plot
# plot_circos(sub, c(2, 4), "Circos Plot.pdf", 8.5, 8.5)
plot_circos(sub, c(2, 15), "Circos Plot with Labels.pdf", 18, 19, TRUE)

```

# Plot Chord Graph

First, transform network to adjacency matrix, then group by function. We also define a utility `merge` function to retrieve the group for an arbitrary set of symbols - this will be useful later.

```{r transform-adjacency}

# utility function to retrieve group
get_grp = function(x, var) {
  merge(x, dat[, .(Symbol, Group)], by.x = var, by.y = "Symbol", all.x = TRUE, sort = FALSE)
}

# melt to adjacency list
adj = as.matrix(as_adj(sub)) %>% reshape2::melt() %>% setDT()

# replace protein names with groups, then convert back to matrix
adj = adj %>% .[value == 1, ] %>% get_grp("Var1") %>% get_grp("Var2") %>%
  .[, .(Group.x, Group.y)] %>% graph_from_data_frame() %>% as_adj(sparse = FALSE)

# reorder alphabetically to match colors
adj = names(cols) %>% adj[., .]

```

The package `circlize` is used to create static chord diagrams, while the GitHub package [`mattflor/chorddiag`](https://github.com/mattflor/chorddiag) is used to create interactive chord diagrams using the Javascript visualization library D3. Notice that, for the static plot, the alternating order defined previously (in `alt`) is used.

```{r plot-chord}

# static plot
pdf(file.path(dir1, "Chord Diagram.pdf"), 12, 12)
# circlize::chordDiagram(adj[alt, alt], transparency = 0.5, grid.col = cols[alt])
circlize::chordDiagram(adj, transparency = 0.5, grid.col = cols)
dev.off()

# interactive plot
chorddiag(adj, groupColors = as.character(cols), groupnamePadding = 30, groupnameFontsize = "10") %>%
  htmlwidgets::saveWidget(., file.path(dir1, "Interactive Chord Diagram.html"))

```

# Plot Network Graph

Create manual grouped layout to cluster by functional group (inspired by `igraph::crossing()` and [this](https://stackoverflow.com/questions/16390221/how-to-make-grouped-layout-in-igraph) StackOverflow post).

```{r group-layout}

# check if two nodes are within the same cluster
el = as_edgelist(sub) %>% as.data.table() %>% 
  get_grp("V1") %>% get_grp("V2") %>%
  .[, Weights := ifelse(Group.x == Group.y, yes = 5, no = 25)]

# # if nodes have no connections with their cluster, set weight as intermediate (to avoid clusters of n = 1)
# for(v in V(sub)$name) {
#   if(el[V1 == v | V2 == v, sum(Weights == 5) <= 1]) { el[V1 == v | V2 == v, Weights := 15] }
# }

# # create manual layout
lyt = graphlayouts::layout_with_stress(sub, weights = el[, Weights]) %>%
  as.data.table() %>% setnames(c("X", "Y")) %>% .[, Symbol := vertex_attr(sub, "name")]

# create manual layout
# lyt = igraph::layout_with_kk(sub, weights = el[, Weights]) %>%
#   as.data.table() %>% setnames(c("X", "Y")) %>% .[, Symbol := vertex_attr(sub, "name")]

```

Then, create network graph.

```{r network-graph}

# create network graph
network_graph = plot_network(sub, "manual", c(1, 10), x = lyt[, X], y = lyt[, Y], arc = FALSE) +
  geom_node_label(aes(label = name), repel = TRUE, alpha = 0.8, 
                  box.padding = 0.5, segment.size = NA, label.size = 0.1, size = 8*(5/14)) +
  theme(legend.title = element_text(size = 16, face = "bold"),
        legend.text = element_text(size = 12.5), legend.position = "bottom")

# save network graph
ggsave(file.path(dir1, "Network Graph.pdf"), network_graph, width = 15, height = 15)

```

Finally, prune the network graph for the methods figure visualization.

```{r methods-network}

# prune network graph
prune_graph = sub %>% delete_vertices(which(igraph::degree(.) <= 25))
prune_lyt = lyt[Symbol %in% vertex_attr(prune_graph, "name")]
  
# create methods graph
methods_network = plot_network(prune_graph, "manual", c(5, 15), x = prune_lyt[, X], y = prune_lyt[, Y], arc = FALSE) + theme(legend.position = "none")

# save methods graph
ggsave(file.path(dir1, "Methods Graph.svg"), methods_network, width = 10, height = 10)

```

# Hub Gene Networks

Create networks for top hub markers emphasizing their connectivity. First, define a function

```{r hub-network}

# remove and recreate directory if it exists
hub_dir = file.path(dir1, "Hub Gene Networks")
if(fs::dir_exists(hub_dir)) fs::dir_delete(hub_dir); fs::dir_create(hub_dir)

hub_network = function(hub, idx, PDF = TRUE, SVG = TRUE) {
  
  # duplicate network, identify hub node index
  hub_net = copy(sub)
  hub_idx = vertex_attr(hub_net, "name") == hub
  
  # label nodes by neighbor/hub status
  # cat(paste0("[", idx, "] ", hub, ", "))
  vertex_attr(hub_net, "hub_node") = V(hub_net) %in% ego(hub_net, order = 1, nodes = hub)[[1]] %>%
    as.numeric() %>% data.table(Label = .) %>%
    .[hub_idx, Label := 2] %>%
    .[, Label := factor(Label, levels = c(0, 1, 2), labels = c("node", "neighbor", "hub"))] %>%
    .[, as.character(Label)]
  
  # identify edges connected to the hub node
  edge_attr(hub_net, "hub_edge") = as_edgelist(hub_net) %>% data.table() %>%
    pmap_chr(paste) %>% grepl(hub, .)
  
  # create sub network for hub nodes and edges
  sub_node_idx = which(vertex_attr(hub_net, "hub_node") %in% c("node", "hub"))
  sub_node_net = delete_vertices(hub_net, sub_node_idx)
  sub_edge_net = delete_edges(hub_net, which(!edge_attr(hub_net, "hub_edge")))
  
  # plot hub network
  hub_graph = ggraph(hub_net, "manual", x = lyt[, X], y = lyt[, Y]) +
    geom_edge_link0(aes(width = hub_edge, color = hub_edge, alpha = hub_edge)) +
    scale_edge_width_manual(values = c(0.2, 0.25)) +
    scale_edge_color_manual(values = c("#3D405B", "#D3D3DA")) +
    scale_edge_alpha_manual(values = c(0.6, 0.4)) +
    geom_node_point(aes(color = hub_node, size = hub_node)) + scale_size_manual(values = c(0, 2, 1)) +
    scale_color_manual(values = c("#DB876B", "#FED766", "#3D405B")) +
    
    # replot hub edges, then nodes, on top of previous graph (so they are non-overlapping)
    geom_edge_link0(data = get_edges()(create_layout(sub_edge_net, layout = "manual", x = lyt[, X], y = lyt[, Y])),
                    width = 0.25, color = "#D3D3DA", alpha = 0.6) +
    geom_node_point(data= get_nodes()(create_layout(sub_node_net, layout = "manual",
                                                    x = lyt[-sub_node_idx, X], y = lyt[-sub_node_idx, Y])),
                    color = "#FED766", size = 2) +
    
    # plot hub node with label
    ggforce::geom_circle(aes(x0 = lyt[hub_idx, X], y0 = lyt[hub_idx, Y], r = 5*(5/14)), fill = "#DB876B", color = NA) +
    geom_text(aes(x = lyt[hub_idx, X], y = lyt[hub_idx, Y]),
              label = hub, size = (3 + 4 * 1/nchar(hub))*(5/14), fontface = "bold", color = "white") +
    theme_graph(fg_text_colour = "white", base_family = "Helvetica") +
    theme(legend.position = "none")
  
  # save output
  if(PDF) {ggsave(file.path(hub_dir, paste(idx, "-", hub, "Network.pdf")), hub_graph, width = 7, height = 7)}

  if(SVG) {ggsave(file.path(hub_dir, paste0(idx, "_", hub, "_Network.svg")),
                  hub_graph + theme(panel.background = element_rect(fill = "#2B2F48", color = "#2B2F48"),
                                    plot.background = element_rect(fill = "#2B2F48", color = "#2B2F48")),
                  width = 7, height = 7)}

  return(hub_graph)
  
}

# map function over all markers
hub_graphs = imap(dat[Symbol %in% vertex_attr(sub, "name"), Symbol], ~hub_network(.x, .y, TRUE, TRUE))
cat("DONE")

```

# Interactive Network Graph

Create interactive network graph (with manual layout identical to above).

```{r interactive-network}

# get network data
net_dat = get.data.frame(sub, "both")

# parse nodes for interactive graph
nodes = net_dat$vertices %>% as.data.table() %>%
  setnames(c("name", "Centrality"), c("id", "value")) %>%
  .[, label := id] %>%
  .[, value := 60*(value+0.5)] %>%
  .[, shape := "circle"] %>%
  .[, color := factor(Group, levels = names(cols), labels = cols)] %>%
  .[, .(id, label, value, shape, color)]

# parse edges for interactive graph
edges = net_dat$edges %>% as.data.table() %>%
  setnames("escore", "value") %>%
  .[, value := value + 0.1] %>%
  .[, color := "rgba(0,0,0,0.5)"] %>%
  .[, title := paste(from, to, sep = " to ")] %>%
  .[, .(from, to, value, color, title)]

# create interactive network graph
int_net = visNetwork(nodes, edges) %>%
  visOptions(height = 800, width = 1200,
             highlightNearest = TRUE, clickToUse = TRUE,
             nodesIdSelection = list(main = "Select Gene")) %>%
  visInteraction(navigationButtons = TRUE, dragView = FALSE) %>%
  visIgraphLayout("layout_with_kk",
                  coords = as.matrix(lyt[, .(X, Y)]), maxiter = 0)

visSave(int_net, file.path(dir1, "Interactive Network Graph.html"))

```
