---
title: "Transcription Factor Enrichment Analysis"
description: |
 This script performs transcription factor analyses using TFEA.ChIP and Enrichr.
bibliography: astrocyte-review-bibliography.bib
csl: https://www.zotero.org/styles/elsevier-vancouver
author:
  - first_name: "Ayush"
    last_name: "Noori"
    url: https://www.github.com/ayushnoori
    affiliation: Massachusetts General Hospital
    affiliation_url: https://www.serranopozolab.com
    orcid_id: 0000-0003-1420-1236
output:
  distill::distill_article:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Dependencies

Load requisite packages. This script uses my personal utilities package `brainstorm`, which can be downloaded via `devtools::install_github("ayushnoori/brainstorm")`.

```{r load-packages, message=FALSE, warning=FALSE}

# data manipulation
library(data.table)
library(purrr)
library(magrittr)

# TFEA package
library(TFEA.ChIP)

# access Enrichr API
library(httr)
library(rjson)

# Excel manipulation
library(openxlsx)

# string manipulation
library(stringr)

# data visualization
library(ggplot2)
library(ggpubr)
library(plotly)
library(htmlwidgets)
library(RColorBrewer)

# utility functions
library(brainstorm)

```

Note that directories are relative to the R project path.

```{r define-directores}

# set directories
ddir = file.path("Data", "3 - Transcription Factor Enrichment Analysis")
dir3 = file.path("Results", "3 - Transcription Factor Enrichment Analysis")

```

# Read Data

Read ADRA protein set and load ChIP-Seq experiment data [@lachmann_chea_2010; @dunham_integrated_2012; @fishilevich_genehancer_2017; @cheneby_remap_2018].

```{r read-data}

# read data
dat = fread(file.path("Data", "ADRA Protein Set.csv"))

# load ChIP-Seq experiment data
load(file.path(ddir, "ReMap+GH_doubleElite.Rdata"))
set_user_data(MetaData, Mat01)

```

# `TFEA.ChIP` Analysis

Perform Transcription Factor Enrichment Analysis (TFEA) using the `TFEA.ChIP` package [@puente-santamaria_tfeachip_2019]. As a control list is not provided, use all human genes NOT in the ADRA marker set as the control.

```{r tfea-analysis}

# convert ADRA gene symbols to ENTREZ
entrez = GeneID2entrez(gene.IDs = dat[, Symbol])

# compute contingency matrices
CM = contingency_matrix(entrez)
pval = getCMstats(CM) %>% as.data.table()

# compute confidence intervals
CI = map(CM, ~fisher.test(.x, conf.int = TRUE, conf.level = 0.95)$conf.int) %>%
  { data.table(Accession = names(.), CI.left = map(., 1), CI.right = map(., 2)) }

# merge with confidence intervals
pval = merge(pval, CI, by = "Accession", all.x = TRUE)[order(-abs(distance)), ]

# write to Excel file
write.xlsx(pval, file.path(dir3, "TFEA Association Test.xlsx"), asTable = TRUE, tableStyle = "TableStyleMedium19")

```

# `TFEA.ChIP` Volcano Plot

Select transcription factors to highlight and define colors.

```{r select-tfs}

# identify top TFs
topTFs = summary(as.factor(pval[1:200, TF]))
print(topTFs[order(-topTFs)][1:5])

# select top TFs
TF = c("CTCF", "ESR1")
names(TF) = TF

# select plotting colors
TFcol = c("#00FFD9", "#FF006F")

```

Create interactive volcano plot of transcription factor enrichment analysis results.

```{r plot-tfs}

# create volcano plot
CMplot = plot_CM(pval, plot_title = "", specialTF = TF, TF_colors = TFcol)

# create axis sequences
xseq = pval[, log10.adj.pVal] %>% { seq(min(., na.rm = TRUE), max(., na.rm = TRUE), by = 0.1) }
yseq = pval[, log2.OR] %>% { seq(min(., na.rm = TRUE), max(., na.rm = TRUE), by = 0.1) }

# edit volcano plot
CMplot = CMplot %>%
  
  # add p-value threshold line
  add_lines(x = -log10(0.05), y = yseq, line = list(dash="dash", width = 1.5, color="red"), inherit = FALSE, hoverinfo = "none", name = "P-Value < 0.05", visible = "legendonly") %>%
  
  # add OR threshold lines
  add_lines(x = xseq, y = log2(1.5), line = list(dash="dash", width = 1.5, color="blue"), inherit = FALSE, hoverinfo = "none", name = "OR > 1.5", visible = "legendonly") %>%
add_lines(x = xseq, y = -log2(1.5), line = list(dash="dash", width = 1.5, color="blue"), inherit = FALSE, hoverinfo = "none", name = "OR < -1.5", visible = "legendonly") %>%
  
  # add axis labels
layout(xaxis = list(title = "-Log<sub>10</sub> Adjusted P-Value"), yaxis = list(title = "Log<sub>2</sub> Odds Ratio"))  %>%
  
  # configure SVG output
  config(toImageButtonOptions = list(format = "svg", filename = "TFEA Plot SVG", width = 2000, height = 1200))

# save figure
htmlwidgets::saveWidget(CMplot, file.path(dir3, "TFEA Plot.html"))
saveRDS(CMplot, file.path(dir3, "TFEA Plotly Object.rds"))

```

Create enhanced, static volcano plot of transcription factor enrichment analysis results.

```{r plot-volcano}

# prepare data
pval_vp = copy(pval) %>%
  .[, Highlight := TF] %>%
  .[!(Highlight %in% c("CTCF", "ESR1")), Highlight := "Other"]

# plot data
vp = ggplot(pval_vp, aes(x = log2.OR, y = log10.adj.pVal, fill = Highlight)) +
  geom_hline(yintercept = 0, color = "black", size = 0.3) +
  geom_vline(xintercept = -1.99, color = "black", size = 0.3) +
  geom_point(alpha = 0.3, size = 1.2, shape = 21, stroke = 0) +
  scale_fill_manual("", values = c("#01CB5F", "#F17F29", "#687278")) +
  geom_point(data = pval[TF == "CTCF"], fill = "#01CB5F", size = 1.2, shape = 21, stroke = 0) +
  geom_point(data = pval[TF == "ESR1"], fill = "#F17F29", size = 1.2, shape = 21, stroke = 0) +
  geom_hline(yintercept = -log10(0.05), linetype="dashed", color = "#EF476F", size = 0.3) +
  geom_vline(xintercept = -log2(1.5), linetype="dashed", color = "#3066BE", size = 0.3) +
  geom_vline(xintercept = log2(1.5), linetype="dashed", color = "#3066BE", size = 0.3) +
  coord_cartesian(xlim = c(-1.99, 1.99), ylim = c(-0.01, 1.99), expand = FALSE) +
  xlab(bquote(bold(log[2]*' Fold-Change'))) +
  ylab(bquote(bold(-log[10]*' FDR '*bolditalic(p)*'-value'))) +
  theme_light() +
  theme(legend.text = element_text(size = 10, face = "bold"),
        legend.position = c(0.1, 0.93),
        legend.background = element_rect(fill = NA, color = NA),
        axis.ticks = element_line(color = "black", size = 0.3),
        panel.border = element_blank())

# save plot
# ggsave(file.path(dir3, "TFEA Volcano Plot.pdf"), vp, width = 8, height = 5.5)
ggsave(file.path(dir3, "TFEA Volcano Plot.svg"), vp, width = 8, height = 5.5)
ggsave(file.path(dir3, "TFEA Volcano Plot.pdf"), vp, width = 8, height = 5.5)

```

# `Enrichr` Analysis

Perform TFEA using the `Enrichr` API [@kuleshov_enrichr_2016].

```{r define-parameters}

# define API call parameters
my_genes = gsub("^\\s+|\\s+$", "", dat[, Symbol])
my_library = "ENCODE_and_ChEA_Consensus_TFs_from_ChIP-X"
my_description = "ADRA Marker Set"

```

Function for API call. This function was adapted from the Python scripts provided in the Enrichr API documentation [here](https://maayanlab.cloud/Enrichr/help#api).

```{r enrichr-api}

enrichr_API = function(genes, description, library) {
  
  # create API call to add gene list
  add_url = "http://amp.pharm.mssm.edu/Enrichr/addList"
  genes_str = paste(genes, collapse="\n")
  payload = list(list = genes_str, description = description)
  
  # make and parse API call
  add_request = httr::POST(url = add_url, body = payload)
  add_response = httr::content(add_request, as = "text", encoding = "UTF-8") %>% fromJSON()
  
  # wait before next call
  Sys.sleep(0.5)
  
  # create URL for next API call to calculate enrichment
  enrich_url = paste0("http://amp.pharm.mssm.edu/Enrichr/enrich", "?userListId=",
                      add_response$userListId, "&backgroundType=", library)
  
  # make and parse next API call
  request = httr::GET(url = enrich_url)
  response = httr::content(request, as = "text", encoding = "UTF-8") %>% fromJSON() %>% .[[1]]
  
  # convert to data.table
  convDT = function(res) { res[[6]] = toString(res[[6]]); return(res)  }
  response = map(response, convDT) %>% rbindlist()
  setnames(response, c("Rank", "Term", "p.value", "z.score", "combined.score", "Genes", "adj.p.value", "old.p.value", "old.adj.p.value"))
  
  # calculate -log10(adj. p-value), then filter
  response[, log.adj.p := -log10(adj.p.value)]
  response = response[, .(Term, p.value, adj.p.value, log.adj.p, z.score, combined.score, Genes)]
  
  # return response object
  return(response)

}

```

Make API call to calculate enrichment.

```{r enrichr-analysis}

# make API call
enrichr = enrichr_API(my_genes, my_description, my_library)

# write to Excel file
write.xlsx(enrichr, file.path(dir3, "Enrichr Results.xlsx"), asTable = TRUE, tableStyle = "TableStyleMedium19")

```

# `Enrichr` Plot

Function to plot `Enrichr` results.

```{r plot-enrichr}

plot_enrichr = function(enrichr_results, library = NULL) {
  
  # for TFEA only
  enrichr_results[, Term := word(Term, 1)]
  
  # select top TFs
  enrichr_results = enrichr_results[order(-log.adj.p), ][1:10, ]
  enrichr_results[, Term := factor(Term, levels = rev(Term))]
  
  # create barplot
  p = ggplot(enrichr_results, aes(x = Term, y = log.adj.p, fill = log.adj.p))+
    geom_col() +
    coord_flip() +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
    geom_text(x = 1.7, y = -log10(0.05) + 0.03, label = "italic(p)*'-value < 0.05'", angle = 270, size = 2.5, parse = TRUE) +
    scale_fill_gradient(low="#FFD166", high="#A63446") +
    scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
    theme_light() +
    theme(plot.title = element_text(size = 16, color = "black", face = "bold", hjust = 0.5),
          axis.title.x = element_blank(), axis.title.y = element_blank(),
          legend.position = "none",
          strip.text = element_text(size=12, color = "black", face="bold"),
          strip.background = element_rect(color=NA, fill="#D9D9D9", size=1, linetype="solid"))
  
  if(!is.null(library)) {
    enrichr_results[, Library := library]
    p = p + facet_wrap(Library ~ ., scales = "free", ncol = 1)
  } else {
    p = p + ylab(bquote(bold(-log[10]*'(adj. '*bolditalic(p)*'-value)'))) + theme(axis.title.x = element_text(size = 14, color = "black"))
  }
  
  return(p)

}

```

Save plots for multiple `Enrichr` libraries. To create plots of multiple `Enrichr` results, use `map` as `plot_enrichr(enrichr, library_name)`, then use `ggarrange(plotlist = ., ncol = 1, align = "hv")` to create composite figure.

```{r multiple-enrichr}

cplots = plot_enrichr(enrichr) # %>% annotate_figure(bottom = text_grob(bquote(bold(-log[10]*'(adj. '*bolditalic(p)*'-value)')), size = 14, color = "black", hjust = 0))

# save figure
ggsave(file.path(dir3, "Enrichr Barplot.pdf"), cplots, height = 4, width = 12)

```
